---
title: "Identifying Spam Tests Through Machine Learning"
author: "KJ Monahan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require(tidyverse))
  install.packages('tidyverse')
library(tidyverse)
if (!require(caret))
  install.packages('caret')
library(caret)
if (!require(tidytext))
  install.packages('tidytext')
library(tidytext)
if (!require(tokenizers))
  install.packages('tokenizers')
library(tokenizers)
if (!require(textutils))
  install.packages('textutils')
library(textutils)
if (!require(caret))
  install.packages('caret')
library(caret)
```

## Introduction

Your phone dings. You check your texts, expecting to see a message from a friend
or family member, but instead, you see an ad. Spam text messages remain a common
annoyance. A 2025 survey by Orbit Media found that 29% of respondents reported
receiving spam messages daily or multiple times a day, with another 28% reporting
receiving them weekly. Spam text messages can even be harmful. A Pew Research Center
survey in 2025 found that 61% of U.S. adults reported receiving spam text messages
at least weekly that attempted to get the receiver's personal information.

The SMS Spam Collection dataset offers a set of spam text messages and genuine
text messages ("ham") that can be used to build and test models to identify
spam messages. The dataset consists of two variables: the classification ("spam" or "ham")
and the text of the message. The dataset was released in 2011, and 5,574 messages,
747 of which are spam and 4,827 of which are ham.

One important consideration when comparing models is the cost of a false positive
(identifying a legitimate message as spam) versus the cost of a false negative
(identifying an unwanted message as ham). Text messages can be used for time-sensitive
communication, with the expectation that the receiver will receive an alert immediately.
Real-world examples range from letting someone know you've arrived to two-factor
authentication codes that expire after a short period of time. CTIA, an organization
representing wireless companies in America, also emphasizes the urgency of some
communications, such as flight changes or fraud alerts, in their best practice
guidelines for members. Governmental organizations can use text messaging to
send emergency notifications, such as an active tornado warning or a water main
breakage requiring a temporary boil order. Missing an urgent, or less time-sensitive
but still meaningful message, can cause harm when people rely on text messaging
for communication. CTIA's 2025 annual survey found that Americans exchanged nearly
2.2 trillion text messages in 2025.

However, failing to identify a spam message could be harmful if the message is one of the
many that attempts to defraud the receiver or trick them into handing over 
personal information. The Orbit Media survey found that while ads were still
the most common form of spam text messages, 37% of respondents said that the
spam text messages they received asked for personal information. While the risk
is real, Pew Research Center found that 71% of U.S. adults feel they know at
least a fair amount about how to avoid falling for an online scam (a category that
includes spam text messages). The FTC, among other organizations, provides guidance
on how to recognize and report text message scams, which may help decrease the
likelihood that someone seeing a spam message is actively harmed.

Considering the trade-offs, I determined that recall should be weighted more
heavily than precision--it was more important that someone not miss a real text
message than receive a spam message. To compare models, I calculated the F value,
which combines precision and recall, with a beta of 0.75, weighting precision 0.75 
times that of recall.

**Final summary here**

## Methods and Analysis

### Data Cleaning

The SMSSpamCollection file is a tab-separated text file, with one line per message.
I parsed it into rows and columns using `read_tsv`.
```{r read_and_parse_file}
lines <- readLines("data/SMSSpamCollection")
rows <- str_split(lines, "\t")
messages <- data.frame(parse_factor(map_chr(rows, 1)), parse_character(map_chr(rows, 2)))
colnames(messages) <- c("type", "message")
head(messages, 10) |> tibble()
```

When I tested splitting the messages by words to see what the most popular words
used were, I saw "gt" and "lt" in the list. While "2", "ur", and "4" are all
commonly-known texting abbreviations, especially in 2011, when the dataset was
collected, "gt" and "lt" seemed like odd words to be sent so frequently.

```{r most_popular_words_overall}
messages |>
  unnest_tokens(word, message, to_lower=TRUE, strip_punct=TRUE) |>
  filter(!word %in% stop_words$word) |>
  count(word) |>
  top_n(10, n) |>
  mutate(word = reorder(word, n)) |>
  arrange(desc(n))
```

Reviewing the messages that contain 'lt' or 'gt' indicates that these are
HTML-encoded entities.
```{r review_lt_and_gt}
messages |>
  filter(sapply(message, function(m) {
    words <- list_c(tokenize_words(m, lowercase = TRUE, strip_punct=TRUE))
    any(c('lt', 'gt') %in% words)
  })) |>
  pull(message) |>
  head()
```

Using the `textutils` package, I decoded any encoded HTML entities and stripped
any HTML tags that might still have been included.
```{r clean_html}
messages <- messages |> mutate(message=HTMLrm(HTMLdecode(message)))
messages |>
  unnest_tokens(word, message, to_lower=TRUE, strip_punct=TRUE) |>
  filter(!word %in% stop_words$word) |>
  count(word) |>
  top_n(10, n) |>
  mutate(word = reorder(word, n)) |>
  arrange(desc(n))

```
Now the most popular words all make sense as words that might be used in texts.

Next I standardized phone numbers. Even in the small sample of texts reviewed,
I could see at least one phone number, and I knew that spam messages might instruct
someone to call or a text a certain number. The presence of "call" and "text"
as commonly used words supported that. But since phone numbers can take a variety
of forms, I standardized several known forms to be all digits. This would also
help me later when testing whether the presence of long strings of numbers, such as
phone numbers or account numbers, would be indictative of a spam message.

```{r standardize_phone_numbers}
convert_phone_numbers <- function(s) {
  s |>
    # Handle formats 000-000-0000, 000.000.000, and 000 000 0000
    str_replace_all("(\\d{3})[-,\\.\\s](\\d{3})[-,\\.\\s](\\d{4})",
                    "\\1\\2\\3") |>
    # Handle format (000) 000-0000
    str_replace_all("\\((\\d{3})\\)\\s?(\\d{3})-(\\d{4})", "\\1\\2\\3")
}
messages <- messages |>
  mutate(message = convert_phone_numbers(message))
```


### Analysis

Before doing any analysis, I split the data into training and test sets, so that
I could look for differences between ham and spam text messages in the training
set only. Knowing that, of course, the data set represents a small fraction of
the spam and ham text messages sent, I wanted to use the test set as a stand in
for the unknown or yet-to-be-set text messages that I couldn't analyze.

Given the overall small size of the data set, I wanted to use as much data as
possible for analysis and training. However, the test set still needs to be large
enough to contain meaningful data. A split of 10%-20% of the data for the test
set is common. I chose 20%, so there would be a little over one thousand messages
in the test set. As 2024 study published in npj Digital Media proposed a minimum
data set sizes of N=500-1000 to reduce overfitting, the remaining training set
size of around 4400 messages should be sufficient.

```{r split_data, include=FALSE}
set.seed(2016)
test_index <- createDataPartition(messages$type,
                                  times = 1,
                                  p = 0.2,
                                  list = FALSE)
test_set <- messages[test_index, ]
train_set <- messages[-test_index, ]
```

#### Most Commonly Used Words

I compared the mostly commonly-used words in the training set for ham messages
versus spam messages.

```{r commonly_used_words, echo=FALSE}
train_set_words <- train_set |>
  unnest_tokens(word, message, to_lower=TRUE, strip_punct=TRUE) |>
  filter(!word %in% stop_words$word)

train_set_words |>
  filter(type == 'ham') |>
  count(word) |>
  top_n(10, n) |>
  mutate(word = reorder(word, n)) |>
  arrange(desc(n))

train_set_words |>
  filter(type == 'spam') |>
  count(word) |>
  top_n(10, n) |>
  mutate(word = reorder(word, n)) |>
  arrange(desc(n))
```

There are some apparent differences. "Free" is the second-most-often used words
in spam texts but does not appear in the top 10 at all for ham texts. However,
there is also overlap. "Call", "ur," "2", and "4" appear in both lists. Therefore,
merely identifying words that are used often in spam messages are likely to result
in a lot of false positives. Instead, I calculated the odds ratio for each word
to identify those that were most likely to be in spam messages.

```{r spam_odds_ratio, echo=FALSE}
ham_spam_odds <- train_set_words |>
  count(word, type) |>
  spread(type, n, fill = 0) |>
  mutate(odds_ratio = (spam + 0.5) / (sum(spam) - spam + 0.5) /
           ((ham + 0.5) / (sum(ham) - ham + 0.5))) |>
  arrange(desc(odds_ratio))
ham_spam_odds |>
  head(15) |>
  tibble()
```

This demonstrates that are definitely words that appear more often in spam messages
than in ham messages. There are 7,513 unique words used in the training set.
2,087 of those words have an odds ratio greater than 1, indicating a positive
correlation between the presence of that word in the message and the message
being spam. 759 of the words have an odds ratio greater than 10, indicating an
even stronger positive correlation.

I then counted the number of times a word in the list of top spam words (those 
that are at least 10 times more likely to appear in a spam message than a ham
message in the training set) and added that data to both the training and test
sets.

```{r count_spam_words}
# Get the words with an odds ratio of 10 or more. That's the list of words
# that are at least 10x more likely to be in a spam message than a ham message.
top_spam_words <- ham_spam_odds |> filter(odds_ratio >= 10) |> pull('word')

# Count how many times any of those words appear in the message
count_spam_words <- function(data, word_list = top_spam_words) {
  data |>
    mutate(top_50_word_count = sapply(message, function(m) {
      words <- list_c(tokenize_words(m, lowercase = TRUE, strip_punct=TRUE))
      sum(words %in% top_50_spam_words)
    }))
}

train_set <- count_spam_words(train_set)
test_set <- count_spam_words(test_set)
```

#### Numbers

Previously I'd hypothesized that spam messages might be more likely to contain
phone numbers or other long number strings, such as a (faux) bank account or
package tracking number. In addition, the list of words more likely to be in
spam messages contained several numbers.

First, I added two counts to my data sets: the total number of digits appearing
in the message and the length of the longest unbroken string of numbers.

```{r count_numbers}
add_number_columns <- function(data) {
  data |>
    mutate(message = convert_phone_numbers(message)) |>
    mutate(num_count = str_count(message, "\\d")) |>
    mutate(longest_number = unlist(lapply(str_match_all(message, "\\d+"), function(m) {
      if (length(m) == 0) {
        0
      } else {
        max(nchar(m[,1]))
      }
    })))
}
train_set <- add_number_columns(train_set)
test_set <- add_number_columns(test_set)
```

Spam messages tend to have both more digits overall and a longer string of
digits than ham messages do.

```{r plot_number_comparison, echo=FALSE}
# Box plot comparing total number of digits
train_set |> 
  ggplot(aes(type, num_count, fill=type)) +
  xlab("Message type") +
  ylab("Number of digits in message") +
  ggtitle("Comparison of digits per message by message type") +
  geom_boxplot()

# Box plot to compare longest string of numbers by type of message
train_set |> 
  ggplot(aes(type, longest_number, fill=type)) + 
  xlab("Message type") +
  ylab("Longest digits in message") +
  ggtitle("Comparison of lengths of longest string of digits per message by message type") +
  geom_boxplot()
```

#### Capital Letters

Like numbers, it's possible that capital letters could differ between spam and
ham messages. However, with few options to convey tone over text messages (the
dataset was published in 2011, which predates the global rise in emoji popularity
from 2014 onward), capital letters or entire words or messages in all caps might
be used by both spam messages and ham messages to convey excitement or urgency.

To see if there was any difference, I first counted both the total number of capital
letters and the longest string that did not contain lowercase letters or digits.
I did allow for white space and punctuation in order to detect the difference
between a single word in all-caps and an entire message in all-caps, so long as
the string started and ended with a capital letter (so removing trailing white
space and punctuation).

```{r capital_letter_columns}
add_capital_columns <- function(data) {
  data |>
    mutate(capital_count = str_count(message, "[A-Z]")) |>
    mutate(longest_capitals = unlist(lapply(str_match_all(message, "[A-Z][^a-z\\d]*[A-Z]"), function(m) {
      if (length(m) == 0) {
        0
      } else {
        max(nchar(m[,1]))
      }
    })))
}

train_set <- add_capital_columns(train_set)
test_set <- add_capital_columns(test_set)
```

On average, spam messages contained more capital letters than ham messages.
However, ham messages show a number of outliers.
```{r compare_number_of_caps, echo=FALSE}
train_set |> 
  ggplot(aes(type, capital_count, fill=type)) + 
  xlab("Message type") +
  ylab("Number of capital letters (A-Z)") +
  ggtitle("Comparison of number of capital letters per message by message type") +
  geom_boxplot()
```

# https://www.pewresearch.org/internet/2025/07/31/online-scams-and-attacks-in-america-today/
# https://www.orbitmedia.com/blog/spam-statistics/
# https://api.ctia.org/wp-content/uploads/2023/05/230523-CTIA-Messaging-Principles-and-Best-Practices-FINAL.pdf
# https://consumer.ftc.gov/articles/how-recognize-report-spam-text-messages
# https://api.ctia.org/wp-content/uploads/2025/10/Messaging-Security-Best-Practices-_October-2025.pdf
# https://dl.acm.org/doi/epdf/10.1145/2034691.2034742
# https://rafalab.dfci.harvard.edu/dsbook-part-2/ml/resampling-methods.html
# https://www.nature.com/articles/s41746-024-01360-w
# https://en.wikipedia.org/wiki/Odds_ratio
# https://blog.emojipedia.org/major-moments-in-emoji-history-1995-to-2025/