---
title: "Identifying Spam Tests Through Machine Learning"
author: "KJ Monahan"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
references:
  - type: report
    id: PewResearchCenter2025
    author:
      - family: Gottfried
        given: Jeffrey
      - family: Park
        given: Eugenie
      - family: Anderson
        given: Monica
    title: "Online Scams and Attacks in America Today"
    container-title: "Pew Research Center"
    URL: https://www.pewresearch.org/internet/2025/07/31/online-scams-and-attacks-in-america-today/
    issued:
      date-parts:
        - - 2025
          - 7
          - 31
    accessed:
      date-parts:
        - - 2026
          - 3
          - 15
  - type: post-weblog
    id: OrbitMedia2025
    author:
      - family: Crestodina
        given: Andy
    title: "2025 Spam Report: The Abuse of Trust in Email and Social (now with AI)"
    container-title: "Orbit Media Studios"
    URL: https://www.orbitmedia.com/blog/spam-statistics/
    issued:
      date-parts:
        - - 2025
          - 12
          - 2
    accessed:
      date-parts:
        - - 2026
          - 3
          - 15
  - type: standard
    id: CTIAMessaging
    author: "CTIA"
    title: "Messaging Principles and Best Practices"
    URL: https://api.ctia.org/wp-content/uploads/2023/05/230523-CTIA-Messaging-Principles-and-Best-Practices-FINAL.pdf
    issued:
      date-parts:
        - - 2023
          - 5
    accessed:
      date-parts:
        - - 2026
          - 3
          - 15
  - type: standard
    id: CTIASecurity
    author: "CTIA"
    title: "Messaging Security Best Practices"
    URL: https://api.ctia.org/wp-content/uploads/2025/10/Messaging-Security-Best-Practices-_October-2025.pdf
    issued:
      date-parts:
        - - 2025
          - 10
    accessed:
      date-parts:
        - - 2026
          - 3
          - 15
  - type: webpage
    id: FTC
    author: "Federal Trade Commission"
    container-title: "Federal Trade Commission Consumer Advice"
    URL: https://consumer.ftc.gov/articles/how-recognize-report-spam-text-messages
    issued:
      date-parts:
        - - 2022
          - 7
    accessed:
      date-parts:
        - - 2026
          - 3
          - 15
  - type: report
    id: CTIAAnnualSurvey
    author: "CTIA"
    title: "2025 Annual Survey Highlights"
    URL: https://api.ctia.org/wp-content/uploads/2025/09/2025-Annual-Survey-Highlights.pdf
    issued:
      date-parts:
        - - 2025
          - 9
          - 8
    accessed:
      date-parts:
        - - 2026
          - 3
          - 15
  - type: paper-conference
    id: Almeida2011
    container-title: "DocEng '11: Proceedings of the 11th ACM symposium on Document engineering"
    publisher: "Association for Computing Machinery"
    title: "Contributions to the Study of SMS Spam Filtering:New Collection and Results"
    author:
      - family: Almeida
        given: Tiago
      - family: María Gómez Hidalgo
        given: José
      - family: Yamakami
        given: Akebo
    URL: https://dl.acm.org/doi/epdf/10.1145/2034691.2034742
    published:
      date-parts:
        - - 2011
          - 9
          - 19
    accessed:
      date-parts:
        - - 2026
          - 3
          - 15
    page: 259-262
  - type: dataset
    id: SMSDataset
    author:
      - family: Almeida
        given: Tiago
      - family: María Gómez Hidalgo
        given: José
    title: "SMS Spam Collection"
    publisher: "UCI Machine Learning Repository"
    DOI: https://doi.org/10.24432/C5CC84
    URL: https://archive.ics.uci.edu/dataset/228/sms+spam+collection
    issued: 2011
    accessed:
      date-parts:
        - - 2026
          - 3
          - 9
  - type: book
    medium: Web
    id: Irizarry2
    edition: 2
    title: "Introduction to Data Science: Statistics and Prediction Algorithms Through Case Studies"
    URL: https://rafalab.dfci.harvard.edu/dsbook-part-2/
    author:
      - family: Irizarry
        given: Rafael
  - type: article-journal
    id: Zantvoort2024
    title: "Estimation of minimal data sets sizes for machine learning predictions in digital mental health interventions"
    container-title: "npj Digital Medicine"
    author:
      - family: Zantvoort
        given: Kirsten
      - family: Nacke
        given: Barbara
      - family: Görlich
        given: Dennis
      - family: Hornstein
        given: Silvan
      - family: Jacobi
        given: Corinna
      - family: Funk
        given: Burkhardt
    DOI: https://doi.org/10.1038/s41746-024-01360-w
    URL: https://www.nature.com/articles/s41746-024-01360-w
    issued:
      date-parts:
        - 2024
        - 12
        - 18
    accessed:
      date-parts:
        - 2026
        - 3
        - 15
  - type: webpage
    id: Bobbitt2021
    title: "How to Interpret Odds Ratios"
    URL: https://www.statology.org/interpret-odds-ratios/
    container-title: Statology
    author:
      - family: Bobbitt
        given: Zach
    issued:
      date-parts:
        - - 2021
          - 8
          - 26
    accessed:
      date-parts:
        - - 2026
          - 3
          - 15
  - type: post-weblog
    id: Broni2025
    title: "Major Moments In Emoji History: 1995* to 2025"
    container-title: "Emojipedia"
    URL: https://blog.emojipedia.org/major-moments-in-emoji-history-1995-to-2025/
    author:
      - family: Broni
        given: Keith
    issued:
      date-parts:
        - - 2025
          - 7
          - 16
    accessed:
      date-parts:
        - - 2026
          - 3
          - 15
  - type: book
    id: Irizarry1
    medium: web
    title: "Introduction to Data Science"
    URL: https://rafalab.dfci.harvard.edu/dsbook/index.html
    edition: 1
    author:
      - family: Irizarry
        given: Rafael
  - type: webpage
    id: GeeksForGeeksKNN
    URL: https://www.geeksforgeeks.org/machine-learning/k-nearest-neighbours/
    container-title: "Geeks for Geeks"
    title: "K-Nearest Neighbor(KNN) Algorithm"
    issued:
      date-parts:
        - - 2026
          - 3
          - 12
    accessed:
      date-parts:
        - - 2026
          - 3
          - 17
  - type: webpage
    id: GeeksForGeeksFBeta
    URL: https://www.geeksforgeeks.org/machine-learning/what-is-f-beta-score/
    container-title: "Geeks for Geeks"
    title: "What is F-Beta Score?"
    issued:
      date-parts:
        - - 2025
          - 7
          - 23
    accessed:
      date-parts:
        - - 2026
          - 3
          - 9
  - type: webpage
    id: Tutorialpedia
    URL: https://www.tutorialpedia.org/blog/check-whether-all-elements-of-a-list-are-in-equal-in-r/
    container-title: "TutorialPedia.org"
    title: "How to Check if All Elements of a List Are Equal in R: Beyond Pairwise identical() – Native Solutions & Custom Functions"
    issued:
      date-parts:
        - - 2025
          - 11
          - 21
    accessed:
      date-parts:
        - - 2026
          - 3
          - 11
  - type: chapter
    id: TidyText
    URL: https://www.tidytextmining.com/tidytext.html
    title: "The tidy text format"
    container-title: "Text Mining with R: A Tidy Approach"
    medium: "Web"
    author:
      - family: Silge
        given: Julia
      - family: Robinson
        given: David
    chapter: 1
    issued:
      date-parts:
        - - 2025
          - 4
          - 7
    accessed:
      date-parts:
        - - 2026
          - 3
          - 12
nocite: |
  @*      
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require(tidyverse))
  install.packages('tidyverse')
library(tidyverse)
if (!require(caret))
  install.packages('caret')
library(caret)
if (!require(tidytext))
  install.packages('tidytext')
library(tidytext)
if (!require(tokenizers))
  install.packages('tokenizers')
library(tokenizers)
if (!require(textutils))
  install.packages('textutils')
library(textutils)
if (!require(caret))
  install.packages('caret')
library(caret)
```

## Introduction

Your phone dings. You check your texts, expecting to see a message from
a friend or family member, but instead, you see an ad. Spam text
messages remain a common annoyance. A 2025 survey by Orbit Media Studios found
that 29% of respondents reported receiving spam messages daily or
multiple times a day, with another 28% reporting receiving them weekly [-@OrbitMedia2025].
Spam text messages can even be harmful. A Pew Research Center survey in
2025 found that 61% of U.S. adults reported receiving spam text messages
at least weekly that attempted to get the receiver's personal
information [-@PewResearchCenter2025].

The SMS Spam Collection dataset [@SMSDataset] offers a set of spam text messages and
genuine text messages ("ham") that can be used to build and test models
to identify spam messages. The dataset consists of two variables: the
classification ("spam" or "ham") and the text of the message. The
dataset was released in 2011, and 5,574 messages, 747 of which are spam
and 4,827 of which are ham [@Almeida2011].

One important consideration when comparing models is the cost of a false
positive (identifying a legitimate message as spam) versus the cost of a
false negative (identifying an unwanted message as ham). Text messages
can be used for time-sensitive communication, with the expectation that
the receiver will receive an alert immediately. Real-world examples
range from letting someone know you've arrived to two-factor
authentication codes that expire after a short period of time. CTIA, an
organization representing wireless companies in America, also emphasizes
the urgency of some communications, such as flight changes or fraud
alerts, in their best practice guidelines for members [-@CTIAMessaging]. Governmental
organizations can use text messaging to send emergency notifications,
such as an active tornado warning or a water main breakage requiring a
temporary boil order. Missing an urgent, or less time-sensitive but
still meaningful message, can cause harm when people rely on text
messaging for communication. CTIA's 2025 annual survey found that
Americans exchanged nearly 2.2 trillion text messages in 2025 [-@CTIAAnnualSurvey].

However, failing to identify a spam message could be harmful if the
message is one of the many that attempts to defraud the receiver or
trick them into handing over personal information. The Orbit Media Studios
survey found that while ads were still the most common form of spam text
messages, 37% of respondents said that the spam text messages they
received asked for personal information [-@OrbitMedia2025]. While the risk is real, Pew
Research Center found that 71% of U.S. adults feel they know at least a
fair amount about how to avoid falling for an online scam (a category
that includes spam text messages) [-@PewResearchCenter2025]. The FTC, among other organizations,
provides guidance on how to recognize and report text message scams [-@FTC],
which may help decrease the likelihood that someone seeing a spam
message is actively harmed.

Considering the trade-offs, I determined that recall should be weighted
more heavily than precision--it was more important that someone not miss
a real text message than receive a spam message. To compare models, I
calculated the F value, which combines precision and recall, with a beta
of 0.75, weighting precision 0.75 times that of recall [@Irizarry1; @GeeksForGeeksFBeta].

**Final summary here**

## Methods and Analysis

### Data Cleaning

The SMSSpamCollection file is a tab-separated text file, with one line
per message. I parsed it into rows and columns using `read_tsv`.

```{r read_and_parse_file}
lines <- readLines("data/SMSSpamCollection")
rows <- str_split(lines, "\t")
messages <- data.frame(parse_factor(map_chr(rows, 1)), parse_character(map_chr(rows, 2)))
colnames(messages) <- c("type", "message")
head(messages, 10) |> tibble()
```

When I tested splitting the messages by words to see what the most
popular words used were, I saw "gt" and "lt" in the list. While "2",
"ur", and "4" are all commonly-known texting abbreviations, especially
in 2011, when the dataset was collected, "gt" and "lt" seemed like odd
words to be sent so frequently.

```{r most_popular_words_overall}
messages |>
  unnest_tokens(word, message, to_lower=TRUE, strip_punct=TRUE) |>
  filter(!word %in% stop_words$word) |>
  count(word) |>
  top_n(10, n) |>
  mutate(word = reorder(word, n)) |>
  arrange(desc(n))
```

Reviewing the messages that contain 'lt' or 'gt' indicates that these
are HTML-encoded entities.

```{r review_lt_and_gt}
messages |>
  filter(sapply(message, function(m) {
    words <- list_c(tokenize_words(m, lowercase = TRUE, strip_punct=TRUE))
    any(c('lt', 'gt') %in% words)
  })) |>
  pull(message) |>
  head()
```

Using the `textutils` package, I decoded any encoded HTML entities and
stripped any HTML tags that might still have been included.

```{r clean_html}
messages <- messages |> mutate(message=HTMLrm(HTMLdecode(message)))
messages |>
  unnest_tokens(word, message, to_lower=TRUE, strip_punct=TRUE) |>
  filter(!word %in% stop_words$word) |>
  count(word) |>
  top_n(10, n) |>
  mutate(word = reorder(word, n)) |>
  arrange(desc(n))

```

Now the most popular words all make sense as words that might be used in
texts.

Next I standardized phone numbers. Even in the small sample of texts
reviewed, I could see at least one phone number, and I knew that spam
messages might instruct someone to call or a text a certain number. The
presence of "call" and "text" as commonly used words supported that. But
since phone numbers can take a variety of forms, I standardized several
known forms to be all digits. This would also help me later when testing
whether the presence of long strings of numbers, such as phone numbers
or account numbers, would be indictative of a spam message.

```{r standardize_phone_numbers}
convert_phone_numbers <- function(s) {
  s |>
    # Handle formats 000-000-0000, 000.000.000, and 000 000 0000
    str_replace_all("(\\d{3})[-,\\.\\s](\\d{3})[-,\\.\\s](\\d{4})",
                    "\\1\\2\\3") |>
    # Handle format (000) 000-0000
    str_replace_all("\\((\\d{3})\\)\\s?(\\d{3})-(\\d{4})", "\\1\\2\\3")
}
messages <- messages |>
  mutate(message = convert_phone_numbers(message))
```

### Analysis

Before doing any analysis, I split the data into training and test sets,
so that I could look for differences between ham and spam text messages
in the training set only. Knowing that, of course, the data set
represents a small fraction of the spam and ham text messages sent, I
wanted to use the test set as a stand in for the unknown or
yet-to-be-set text messages that I couldn't analyze.

Given the overall small size of the data set, I wanted to use as much
data as possible for analysis and training. However, the test set still
needs to be large enough to contain meaningful data. A split of 10%-20%
of the data for the test set is common [@Irizarry2]. I chose 20%, so there would be a
little over one thousand messages in the test set. As 2024 study
 proposed a minimum data set sizes of
N=500-1000 to reduce overfitting [@Zantvoort2024], the remaining training set size of
around 4400 messages should be sufficient.

```{r split_data, include=FALSE}
set.seed(2016)
test_index <- createDataPartition(messages$type,
                                  times = 1,
                                  p = 0.2,
                                  list = FALSE)
test_set <- messages[test_index, ]
train_set <- messages[-test_index, ]
```

#### Most Commonly Used Words

I compared the mostly commonly-used words in the training set for ham
messages versus spam messages.

```{r commonly_used_words, echo=FALSE}
train_set_words <- train_set |>
  unnest_tokens(word, message, to_lower=TRUE, strip_punct=TRUE) |>
  filter(!word %in% stop_words$word)

train_set_words |>
  filter(type == 'ham') |>
  count(word) |>
  top_n(10, n) |>
  mutate(word = reorder(word, n)) |>
  arrange(desc(n))

train_set_words |>
  filter(type == 'spam') |>
  count(word) |>
  top_n(10, n) |>
  mutate(word = reorder(word, n)) |>
  arrange(desc(n))
```

There are some apparent differences. "Free" is the second-most-often
used words in spam texts but does not appear in the top 10 at all for
ham texts. However, there is also overlap. "Call", "ur," "2", and "4"
appear in both lists. Therefore, merely identifying words that are used
often in spam messages are likely to result in a lot of false positives.
Instead, I calculated the odds ratio for each word to identify those
that were most likely to be in spam messages. [@Irizarry1]

```{r spam_odds_ratio, echo=FALSE}
ham_spam_odds <- train_set_words |>
  count(word, type) |>
  spread(type, n, fill = 0) |>
  mutate(odds_ratio = (spam + 0.5) / (sum(spam) - spam + 0.5) /
           ((ham + 0.5) / (sum(ham) - ham + 0.5))) |>
  arrange(desc(odds_ratio))
ham_spam_odds |>
  head(15) |>
  tibble()
```

This demonstrates that are definitely words that appear more often in
spam messages than in ham messages. There are 7,513 unique words used in
the training set. 2,087 of those words have an odds ratio greater than
1, indicating a positive correlation between the presence of that word
in the message and the message being spam. 759 of the words have an odds
ratio greater than 10, indicating an even stronger positive correlation. [@Bobbitt2021]

I then counted the number of times a word in the list of top spam words
(those that are at least 10 times more likely to appear in a spam
message than a ham message in the training set) and added that data to
both the training and test sets.

```{r count_spam_words}
# Get the words with an odds ratio of 10 or more. That's the list of words
# that are at least 10x more likely to be in a spam message than a ham message.
top_spam_words <- ham_spam_odds |> filter(odds_ratio >= 10) |> pull('word')

# Count how many times any of those words appear in the message
count_spam_words <- function(data, word_list = top_spam_words) {
  data |>
    mutate(top_spam_words_count = sapply(message, function(m) {
      words <- list_c(tokenize_words(m, lowercase = TRUE, strip_punct=TRUE))
      sum(words %in% top_spam_words)
    }))
}

train_set <- count_spam_words(train_set)
test_set <- count_spam_words(test_set)
```

#### Numbers

Previously I'd hypothesized that spam messages might be more likely to
contain phone numbers or other long number strings, such as a (faux)
bank account or package tracking number. In addition, the list of words
more likely to be in spam messages contained several numbers.

First, I added two counts to my data sets: the total number of digits
appearing in the message and the length of the longest unbroken string
of numbers.

```{r count_numbers}
add_number_columns <- function(data) {
  data |>
    mutate(message = convert_phone_numbers(message)) |>
    mutate(num_count = str_count(message, "\\d")) |>
    mutate(longest_number = unlist(lapply(str_match_all(message, "\\d+"), function(m) {
      if (length(m) == 0) {
        0
      } else {
        max(nchar(m[,1]))
      }
    })))
}
train_set <- add_number_columns(train_set)
test_set <- add_number_columns(test_set)
```

Spam messages tend to have both more digits overall and a longer string
of digits than ham messages do.

```{r plot_number_comparison, echo=FALSE}
# Box plot comparing total number of digits
train_set |> 
  ggplot(aes(type, num_count, fill=type)) +
  xlab("Message type") +
  ylab("Number of digits in message") +
  ggtitle("Comparison of digits per message by message type") +
  geom_boxplot()

# Box plot to compare longest string of numbers by type of message
train_set |> 
  ggplot(aes(type, longest_number, fill=type)) + 
  xlab("Message type") +
  ylab("Longest digits in message") +
  ggtitle("Comparison of lengths of longest string of digits per message by message type") +
  geom_boxplot()
```

#### Capital Letters

Like numbers, it's possible that capital letters could differ between
spam and ham messages. However, with few options to convey tone over
text messages (the dataset was published in 2011, which predates the
global rise in emoji popularity from 2014 onward [@Broni2025]), capital letters or
entire words or messages in all caps might be used by both spam messages
and ham messages to convey excitement or urgency.

To see if there was any difference, I first counted both the total
number of capital letters and the longest string that did not contain
lowercase letters or digits. I did allow for white space and punctuation
in order to detect the difference between a single word in all-caps and
an entire message in all-caps, so long as the string started and ended
with a capital letter (so removing trailing white space and
punctuation).

```{r capital_letter_columns}
add_capital_columns <- function(data) {
  data |>
    mutate(capital_count = str_count(message, "[A-Z]")) |>
    mutate(longest_capitals = unlist(lapply(str_match_all(message, "[A-Z][^a-z\\d]*[A-Z]"), function(m) {
      if (length(m) == 0) {
        0
      } else {
        max(nchar(m[,1]))
      }
    })))
}

train_set <- add_capital_columns(train_set)
test_set <- add_capital_columns(test_set)
```

Using a ridge plot, I can see that while most ham messages have only a
few capital letters, there's a long tail. There's also overlap between
spam and ham messages at the lower end of the range. Messages with 5 to
30 capitals are more likely to be spam messages, but then spam messages
also have a long tail.

```{r compare_number_of_caps, echo=FALSE}
train_set |> 
  ggplot(aes(capital_count, type, fill=type)) + 
  ylab("Message type") +
  xlab("Number of capital letters (A-Z)") +
  ggtitle("Comparison of total number of capital letters per message by message type") +
  geom_density_ridges(bandwidth = 2, alpha=0.5)
```

Comparing the length of the longest string of capital letters, there's a
similar pattern. Both ham and spam messages have long tails and both
overlap at the bottom of the range, with no string of two or more
capital letters. However, while ham messages are much more likely to
have no string than to have any length of capital letters in a row, spam
messages have a wider distribution with a peak around a string of 5-10
characters.

```{r compare_cap_length, echo=FALSE}
train_set |> 
  ggplot(aes(longest_capitals, type, fill=type)) + 
  xlab("# of messages") +
  ylab("Longest string of capital letters") +
  ggtitle("Comparison of length of the longest string of capital letters per message by message type") +
  geom_density_ridges(bandwidth=2, alpha=0.5)
```

Looking only at messages where `longest_capitals` is greater than 0, I
can see that most of the difference between spam and ham messages is
driven by most ham messages not having a multi-capital-letter string.
However, with those removed, it's more apparent that while spam messages
peak at a slightly greater number, ham messages have a more substantial
tail. This suggests that while having a `longest_capitals` length of 0
might give credence to a message being ham, the length of the string
beyond that doesn't seem to be a clear indicator one way or another.

```{r compare_cap_length_filtered, echo=FALSE}
train_set |> 
  filter(longest_capitals > 0) |>
  ggplot(aes(longest_capitals, type, fill=type)) + 
  xlab("# of messages") +
  ylab("Longest string of capital letters") +
  ggtitle("Comparison using only messages with a capital-letter string") +
  geom_density_ridges(bandwidth=2, alpha=0.5)
```

#### Message length

Finally, I counted the total character length of the messages to examine
if there were any differences.

```{r count_msg_length}
train_set <- train_set |> mutate(msg_length = nchar(message))
test_set <- test_set |> mutate(msg_length = nchar(message))
```

I can see that, on average, ham messages are shorter than spam messages.
However, once again, both types contain a number of outliers, with
messages longer than 250 characters being entirely ham messages.

```{r message_length, echo=FALSE}
train_set |> 
  ggplot(aes(type, msg_length, fill=type)) + 
  xlab("Message type") +
  ylab("Total message length") +
  ggtitle("Comparison of message lengths by message type") +
  geom_boxplot()
```

### Model Development

To compare model performance, I wanted to consider both precision and
recall, but, as discussed earlier, giving more weight to recall. I
tested each model against my previously held-out test set and compared
the F-value, which computes the harmonic average of precision and
recall. I used a weighted harmonic average with a beta value of 0.75, so
that precision was weighted 0.75 times that of recall.

#### Model 1: Guessing

To get a baseline, I first tested an extremely simple model, guessing
whether a message is spam or ham, assuming the same proportions of spam
messages that are in the training set.

```{r model_1}
p_spam <- sum(train_set$type == 'spam') / nrow(train_set)
guess_y_hat <- sample(c('spam', 'ham'), nrow(test_set), replace = TRUE, prob = c(p_spam, 1 - p_spam)) |>
  factor(levels = levels(test_set$type))
guess_f_meas <- F_meas(
  guess_y_hat,
  reference = test_set$type,
  relevant = "spam",
  beta = 0.75
)
guess_f_meas
```

#### Model 2: Using longest number

During the analysis, I found that messages with long strings of digits
are more likely to be spam. For the second model, all messages with the
longest string of digits above a certain cutoff are classified as spam,
all others as ham. To find the cutoff, I compared the F-value against
the training set for cutoffs ranging from 0 to 15. A cutoff of 4
provided the optimal value, so I then tested on the test set, using that
cutoff.

```{r model_2}
# Determined through tuning on the training set
cutoff <- 4
long_number_y_hat <- ifelse(test_set$longest_number >= cutoff, "spam", "ham") |>
  factor(levels = levels(test_set$type))
long_number_f_meas <- F_meas(
  long_number_y_hat,
  reference = test_set$type,
  relevant = "spam",
  beta = 0.75
)
long_number_f_meas
```

This improves performance dramatically over guessing.

#### Model 3: Generalized linear model

Because messages are either spam or ham, which can be represented as 0
and 1, a generalized linear model or logistic regression model makes
more sense than a linear regression model, which would allow for values
below 0 or above 1. A logistic regression model converts the probability
to log odds, calculating how much more likely it is that a message will
be spam than that it will not be spam. Rather than a simple binary
cutoff, I can use a linear regression model to calculate whether a
message is likely spam or ham given a value.

```{r model_3}
glm_digit_length_fit <- train_set |>
  mutate(y = as.numeric(type == "spam")) |>
  glm(y ~ longest_number, data = _, family="binomial")
# Predict the results on the test set, using the model created on the training set
# Returns the likelihood that a message is spam
glm_digit_length_p_hat <- predict(glm_digit_length_fit, newdata=test_set, type="response")
# Translate those probabilities to predictions by saying that any message with
# a greater than 50% chance of being spam is spam
glm_digit_length_y_hat <- ifelse(glm_digit_length_p_hat > 0.5, "spam", "ham") |> 
  factor(levels = levels(test_set$type))
glm_digit_length_f_meas <- F_meas(
  glm_digit_length_y_hat,
  reference = test_set$type,
  relevant = "spam",
  beta = 0.75
)
glm_digit_length_f_meas
```

This did less well than finding a cut-off. While messages with longer
numbers are more likely to be spam, the data is approximately bivariate
normal.

I also considered looking at more than one variable. For example,
messages with more numbers and with more capital letters both tended to
be spam messages.

```{r number_count_vs_capital_count}
train_set |>
  ggplot(aes(num_count, capital_count, color = type)) +
  xlab("Total number of digits") +
  ylab("Total number of capital letters") +
  geom_point()
```

While the data is mostly grouped between ham and spam messages, the
relationship is not linear.

#### Model 4: K-Nearest Neighbor

K-Nearest Neighbor (KNN) identifies the k closest data points to an item
and makes the prediction based on the average of the closest data
points. One advantage it has over logistic or linear regression here is
that it does not assume the data is distributed in any particular way.
Since ham and spam messages tend to be clustered near similar messages,
predicting the classification based on the nearest other messages might
yield better results. [@GeeksForGeeksKNN]

I used the `caret` package to train the model and perform
cross-validation with bootstrap samples to determine which value of k
performed best. Bootstrap sampling takes a sample with replacement of
the training dataset to approximate the full dataset. The `train` method
does this 25 times (the default value) for each parameter and calculates
the accuracy. This approximates the distribution across the full
dataset. I used the same weighted F value for accuracy that has been
used on models previously. Whatever parameter value gives the highest
accuracy when the 25 tests are averaged together is what's used in the
final model.

```{r knn_tune}
knn_fit <- train(
  type ~ ., 
  method = "knn", 
  data = train_set_features, 
  trControl = control, 
  metric = "F1",
  tuneGrid = data.frame(k = c(3, 5, 7, 9, 11))
)
knn_fit$bestTune
```

I then tested the final model against the test set.

```{r knn_test}
knn_y_hat <- predict(knn_fit, newdata = test_set, type="raw")
knn_f_meas <- F_meas(
  knn_y_hat,
  reference = test_set$type,
  relevant = "spam",
  beta = 0.75
)
knn_f_meas
```

#### Model 5: Classification Tree

Another way to classify messages is to look at various combinations of
factors. For instance, I've already seen that messages with long strings
of numbers are likely to be spam. Messages with more than 5 capital
letters in a row are likely spam unless they have more than 50 capital
letters, in which case, they're likely ham.

A classification tree establishes a set of rules or series of if/else
conditions to determine the prediction. The complexity parameter (cp) is
the value that can be tuned here to determine the optimal minimum by
which adding another split must improve the model [@Irizarry2].

```{r class_tree_fit}
class_tree_fit <- train(
  type ~ .,
  method = "rpart",
  data=train_set_features,
  trControl = control,
  metric = "F1",
  tuneGrid = data.frame(cp = seq(0.0, 0.1, len = 25)),
)
class_tree_y_hat <- predict(class_tree_fit, newdata = test_set, type="raw")
class_tree_f_meas <- F_meas(
  class_tree_y_hat,
  reference = test_set$type,
  relevant = "spam",
  beta = 0.75
)
class_tree_f_meas
```

Plotting the final model shows what variables and cut-off values were
used.

```{r class_tree_plot}
plot(class_tree_fit$finalModel, margin = 0.1)
text(class_tree_fit$finalModel, cex = 0.75)
```

The model only needed the number of occurrences of top spam words and
the length of the longest number to predict the outcome.

#### Model 6: Random Forest

Random forest builds on the classification tree idea by randomly
selecting bootstrap samples and building classification trees from the
samples, using randomly selected features. The resulting trees create
the "forest", and the average prediction of all the trees determines the
final prediction [@Irizarry2].

```{r random_tree_fit}
random_forest_fit <- train(
  type ~ .,
  method = "rf",
  data = train_set_features,
  trControl = control,
  metric = "F1",
  # Tune the number of variables randomly assigned per split
  tuneGrid=data.frame(mtry=seq(1, 6))
)
random_forest_y_hat <- predict(random_forest_fit, newdata = test_set, type="raw")
random_forest_f_meas <- F_meas(
  random_forest_y_hat,
  reference = test_set$type,
  relevant = "spam",
  beta = 0.75
)
random_forest_f_meas
```

When reviewing the relative importance of the various factors, I can see
that `top_spam_words_count` had the greatest impact. That is the same
factor that was used in the first split in the classification tree
model.

```{r random_tree_importance, echo=FALSE}
importance(random_forest_fit$finalModel)
```

### Ensembles

After testing models one at a time, I considered whether combining
predictions from multiple models could yield better results.

One challenge is that I wanted to only use models that were among the
most accurate, but I could not use the previously computed accuracy
results, because they were based on test data. Using the results from
testing on the test set to determine which model to use would pollute my
data.

Instead, I used the same bootstrap approach used for tuning the model
parameters. I created five bootstrap samples from the training data,
predicted the outcome for each using the existing models, and averaged
the resulting F-values.

```{r bootstrap_accuracy, echo=FALSE}
folds = createResample(train_set$type, times = 5)
guess_cv <- foreach(fold=folds, .combine = c) %do% {
  partition <- train_set[fold,]
  y_hat <- sample(c('spam', 'ham'), nrow(partition), replace = TRUE, prob = c(p_spam, 1 - p_spam)) |>
    factor(levels = levels(partition$type))
  F_meas(
    y_hat,
    reference = partition$type,
    relevant = "spam",
    beta = 0.75
  )
}
guess_accuracy <- mean(guess_cv)

longest_number_cv <- foreach(fold=folds, .combine = c) %do% {
  partition <- train_set[fold,]
  y_hat <- ifelse(partition$longest_number >= cutoff, "spam", "ham") |>
    factor(levels = levels(partition$type))
  F_meas(
    y_hat,
    reference = partition$type,
    relevant = "spam",
    beta = 0.75
  )
}
longest_number_accuracy <- mean(longest_number_cv)

glm_digit_cv <- foreach(fold=folds, .combine = c) %do% {
  partition <- train_set[fold,]
  p_hat <- predict(glm_digit_length_fit, newdata=partition, type="response")
  y_hat <- ifelse(p_hat > 0.5, "spam", "ham") |> 
    factor(levels = levels(partition$type))
  F_meas(
    y_hat,
    reference = partition$type,
    relevant = "spam",
    beta = 0.75
  )
}
glm_digit_accuracy <- mean(glm_digit_cv)

knn_cv <- foreach(fold=folds, .combine = c) %do% {
  partition <- train_set[fold,]
  y_hat <- predict(knn_fit, newdata = partition, type="raw")
  F_meas(
    y_hat,
    reference = partition$type,
    relevant = "spam",
    beta = 0.75
  )
}
knn_accuracy <- mean(knn_cv)

class_tree_cv <- foreach(fold=folds, .combine = c) %do% {
  partition <- train_set[fold,]
  y_hat <- predict(class_tree_fit, newdata = partition, type="raw")
  F_meas(
    y_hat,
    reference = partition$type,
    relevant = "spam",
    beta = 0.75
  )
}
class_tree_accuracy <- mean(class_tree_cv)

random_forest_cv <- foreach(fold=folds, .combine = c) %do% {
  partition <- train_set[fold,]
  y_hat <- predict(random_forest_fit, newdata = partition, type="raw")
  F_meas(
    y_hat,
    reference = partition$type,
    relevant = "spam",
    beta = 0.75
  )
}
random_forest_accuracy <- mean(random_forest_cv)

tibble(
  Models=c('Guessing', 'Longest number', 'Logistic regression', 'K-Nearest Neighbor', 'Classification Tree', 'Random Forest'),
  `F-Values`=c(guess_accuracy, longest_number_accuracy, glm_digit_accuracy, knn_accuracy, class_tree_accuracy, random_forest_accuracy)
)
```

Five of the models had F-values of greater than 0.9 against the training
data bootstrap samples.

#### Voting

The first ensemble model I tried was a voting method. Since Longest
number and Logistic regression were both models based on the
`longest_number` feature, I only used one of them. I chose Longest
number because it performed better in my bootstrap sample.

This left me with an even number of models, so I had to determine how to
handle ties. Since I had decided that not misidentifying ham was more
important than not misidentifying spam, in case of a tie, the message
was classified as ham. That meant that 3 out of the 4 models had to
consider a message spam in order to classify it as such.

```{r ensemble_voting}
ensemble_vote_fit <- cbind(long_number_y_hat, class_tree_y_hat, random_forest_y_hat, knn_y_hat)
ensemble_vote_y_hat <- apply(ensemble_vote_fit, 1, function(r) {
  ifelse(sum(r == 2) >= 3, 'spam', 'ham')
}) |> factor(levels = levels(test_set$type))
ensemble_vote_f_meas <- F_meas(
  ensemble_vote_y_hat,
  reference = test_set$type,
  relevant = "spam",
  beta = 0.75
)
ensemble_vote_f_meas
```

#### Mean Probability

The second ensemble model I tried was by calculating the probability of
a message being spam, using the top three models: KNN, Classification
Tree, and Random Forest. I took the mean of the three probabilities and
used the mean probability to predict whether the message was spam or
ham.

```{r ensemble_mean}
# Re-generate predictions to get the probability of the item being spam
knn_p_hat <- predict(knn_fit, newdata = test_set , type = "prob")['spam']
class_tree_p_hat <- predict(class_tree_fit, newdata = test_set, type = "prob")['spam']
random_forest_p_hat <- predict(random_forest_fit, newdata = test_set, type = "prob")['spam']
ensemble_mean_fit <- cbind(knn_p_hat, class_tree_p_hat, random_forest_p_hat)
# Get the mean probability across all three models. If it's greater than 0.5
# likelihood of being spam, classify it as such
ensemble_mean_y_hat <- apply(ensemble_mean_fit, 1, function(r) {
  ifelse(mean(r) > 0.5, 'spam', 'ham')
}) |> factor(levels = levels(test_set$type))
ensemble_mean_f_meas <- F_meas(
  ensemble_mean_y_hat,
  reference = test_set$type,
  relevant = "spam",
  beta = 0.75
)
ensemble_mean_f_meas
```

## Results

While the ensemble mean approach improved the results over the
classification tree and KNN models on their own, the random forest model
gave the most accurate results overall.

```{r final_results, echo=FALSE}
tibble(
  Models=c('Guessing', 'Longest number', 'Logistic regression', 'K-Nearest Neighbor', 'Classification Tree', 'Random Forest', 'Ensemble Voting', 'Ensemble Mean'),
  `F-Values`=c(guess_f_meas, long_number_f_meas, glm_digit_length_f_meas, knn_f_meas, class_tree_f_meas, random_forest_f_meas, ensemble_vote_f_meas, ensemble_mean_f_meas)
)
```

The full confusion matrix details that the random forest model has
specificity/recall over 99% and a slightly lower sensitivity/precision.
It correctly classified 99% of the messages in the test set overall and
94% of the spam messages. Importantly, only 2 of the 966 ham messages
were incorrectly classified as spam.

```{r confusion_matrix}
confusionMatrix(random_forest_y_hat, test_set$type, positive = "spam")
```

## Conclusion

Overall, a random forest model works well for classifying spam messages
using factors such as the presence of words known to be used more often
in spam messages, long strings of digits such as phone numbers, and
numbers of digits or capital letters.

Further research could be done using a more recently-collected data set.
The dataset used here was donated in 2011, and changes in technology and
communication styles may have changed what makes a message identifiable
as spam. Other factors that could be considered in a future study would
be whether emoji usage or attachments such as photos are indicative of a
message being spam or ham. In addition, the work here attempts to
identify messages as spam or ham based only on the message content.
Other context about the message itself, such as whether the receiver has
communicated with the sender before or whether the message was sent via
an automated program or another phone, could potentially further improve
classification.
